---
output:
   html_document:
      self_contained: false
      keep_md: true
      preserve_yaml: false
---
<center>

## Harvard Kennedy School
***
### API 201Z: Fall 2020
#### Problem Set #1 - March 26, 2020
#### __Nolin Greene__

</center>

__Question #1: Case Study - Pine Street Inn__

*Data and Package Loading (output masked for readability purposes)*

```{r, loadpackages, message=FALSE}
rm(list=ls())

options(scipen = 999)

library(readxl);library(dplyr);library(ggplot2);library(plotly)
library(tidyr);library(knitr);library(kableExtra);library(stringr)

d<-read_excel("Pine Street Inn Length of Stay Data - Solutions.xls", 
              sheet = 1, cell_cols(1:2))

colnames(d)<-c("n","los")
```
*1.1:* The mean length of stay at Pine Street Inn is **`r round(mean(d$los),0)` days**. 

*1.2:* The median length of stay at Pine Street Inn is **`r round(median(d$los),0)` days**. 

*1.3:* The maximum length of stay at Pine Street Inn is **`r max(d$los)` days** and the minimum length of stay is **`r min(d$los)` day**. 

*1.4:* The 75th percentile length of stay at Pine Street Inn is **`r round(quantile(d$los, probs=.75),0)` days.**  The 95th and 99th percentiles are **`r round(quantile(d$los, probs=.90),0)` days** and **`r round(quantile(d$los, probs=.95),0)` days** respectively

*1.5:* There are **`r sum(d$los)` bednights** represented in the dataset.

*1.6:* There are **`r nrow(d)` guests** represented in the dataset.

*1.7* 
```{r}
d%>%
  filter(los<=3) %>%
  summarize(n=n(), bednights=sum(los))

d%>%
  filter(los<=10 & los>3) %>%
  summarize(n=n(), bednights=sum(los))

d%>%
  filter(los<=35 & los>10) %>%
  summarize(n=n(), bednights=sum(los))

d%>%
  filter(los<=150 & los>35) %>%
  summarize(n=n(), bednights=sum(los))

d%>%
  filter(los>150) %>%
  summarize(n=n(), bednights=sum(los))
```



### Summary Statistics for PSI Length of Stay
```{r, echo = F}
Length_of_Stay<-c("3 Days or Less", "4 to 10 Days", "11 to 35 Days", "36 to 150 Days", "151 Days or More")
Number_of_Guests<-c(721, 1177, 1048, 721, 288)
Number_of_Bed_Nights<-c(4973, 7328, 21007, 53832, 84765)
Fraction_of_Guests<-c(round(721/6556,2), round(1177/6556,2), round(1048/6556,2), round(721/6556,2), round(288/6556,2))
Fraction_of_Bed_Nights<-c(round(4973/sum(d$los),2), round(7328/sum(d$los),2), round(21007/sum(d$los),2), round(53832/sum(d$los),2), round(84765/sum(d$los),2))
AvgBedNightsperGuest<-c(round(4973/3322,0),round(7329/1177,0),round(21007/1048,0),round(53832/721,0),round(84765/288,0))
 
one7<-data.frame(Length_of_Stay,Number_of_Guests,Number_of_Bed_Nights,Fraction_of_Guests,Fraction_of_Bed_Nights,AvgBedNightsperGuest, row.names = 1)
one7<-rbind(one7, Total=c(sum(one7$Number_of_Guests),sum(one7$Number_of_Bed_Nights),"", "", "",""))
names_spaced<-c("Number of <br>Guests", "Number of <br>Bed Nights", "Fraction <br>of Guests", "Fraction of <br>Bed Nights","Avg Bed Nights <br> per Guest")
kable(one7,
      row.names = T,
      col.names = names_spaced, 
      escape = F) %>%
kable_styling(bootstrap_options = c("striped", "hover", "condensed"))

```

*1.8:* 
```{r}
d<-d %>%
  mutate(bin = case_when(
      los<4 ~ "3 Days or Less",
      los>3 & los<11 ~ "4 to 10 Days",
      los>10 & los <36 ~ "11 to 35 Days",
      los>35 & los<151 ~"36 to 150 Days",
      los>150 ~ "151 Days or More"),
    bin = factor(bin, levels = c("3 Days or Less","4 to 10 Days","11 to 35 Days",
                 "36 to 150 Days", "151 Days or More")))

count<-group_by(d, bin)%>%
  summarise(n = n(), los=sum(los))

count<-mutate(count,
       Clients = count$n/sum(count$n),
       Bed_Nights = count$los/sum(count$los))

count<-count[c(1,4,5)]

g<-gather(count, stat, percent, -bin)

p<-ggplot(g,aes(x=bin, y=percent, group = stat, color=stat))+
  geom_line()+
  theme(legend.title = element_blank()) +
  labs(x="Length of Stay", y="Percent of Total", title = "Total Clients and Bednights by Length of Stay")+
  scale_x_discrete(labels = function(x) str_wrap(x, width = 8))

ggplotly(p)
```


*1.9:* 

```{r}
ggplot(d, aes(x=los))+
  geom_histogram(colour="black", fill = "slategray", binwidth = 10)+
  labs(x="Length of Stay (days)", y="Number of Guests", title="Distribution of Guests by Length of Stay")
```

*1.9:* Simply by looking at the mean, one might infer that it is common for a PSI guest to spend 3-4 weeks in shelter. However, upon calculating additional statistics (median, IQR, historgram), we see that the distribution of length of stay is heavily right skewed, with a small number of guests having very long stays. This leads me to believe that Pine Street faces a very severe Pareto Principle, with a small number of guests occupying an extreme proportion of the shelter's total bed stays. 


***

__Question #2: State Spending Data__

*A.* The total direct expenditure was $3.147tr. The total spent on Elementary and Secondary Education was $565bn, the total spent on Health was $84bn and the total spent on Corrections was $72.6bn.

*B.*
```{r, message=FALSE}
temp <- tempfile()
download.file("http://www2.census.gov/govs/local/11statetypepu.zip",temp)
state_exp <- read.table(unz(temp, "11statetypepu.txt"))
colnames(state_exp)<-c("govtype","itemcode","amount", "cv", "yr")
unlink(temp)
```

```{r}
table(state_exp$yr)
state_exp<-subset(state_exp, select = -yr)
state_exp<-filter(state_exp, govtype == 1)
state_exp<-subset(state_exp, select = -govtype)
state_exp$amount<-state_exp$amount/1000
state_exp<-state_exp %>%
  mutate(cat = case_when(
      itemcode=="E32" | itemcode=="F32" | itemcode=="G32" ~ "Health",
      itemcode=="E12" | itemcode=="F12" | itemcode=="G12" ~ "Education",
      itemcode=="E04" | itemcode=="F04" | itemcode=="G04" |
      itemcode=="E05" | itemcode=="F05" | itemcode=="G05" ~ "Corrections"))
state_exp<-filter(state_exp, cat %in% c("Health", "Education", "Corrections"))
state_exp %>%
  group_by(cat)%>%
  summarize(sum = sum(amount))
```

